---
title: "Disproportionality_analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Disproportionality_analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

During the life cycle of a medicinal product, knowledge of the safety profile 
continues to accumulate beyond market introduction. Spontaneous reporting of
adverse events through means of individual case safety reports remains an 
important part of this safety monitoring.

Given a large set of reports covering a multitude of drugs and adverse events, 
disproportionality analysis is a data driven framework to identify 
drug-event-combinations of potential interest for further qualitative 
manual case-by-case assessment. The pvutils package provides three estimators, 
the proportional reporting rate (PRR), the reporting odds ratio (ROR) and the 
information component (IC). 

## Disproportionality analysis for a specific drug-event combination
A first use-case for these functions is to provide a way to calculate 
confidence or credibility intervals for a specific drug-event-combination, where
you already know the number of observed reports, and other required counts such as
the database total count, number of reports with the exposure and number of
reports with the event.

```{r}
library(pvutils)
prr(obs = 10, n_drug = 1000, n_event_prr = 200, n_tot_prr = 10000)
ror(a = 10, b = 20, c = 200, d = 10000)
ic(obs = 10, exp = 5)
```

For further details on the specific input parameters, consult the documentation 
of each function. 

## Disproportionality analysis from report-level data
If the aim is to screen a wide range of drug-event pairs from report-level data 
using disproportionality analysis, this is handled through the function \code{da}.

The \code{pvutils} package contains a simulated data set 'drug_event_df'. For 
simplicity, drugs are named Drug_A - Drug_Z, and events are 
named Event_1 - Event_1000. Further details on the proportions of each drug and
event are described in the data object (execute \code{?df_drug_event}). We pass the 
data to a function that calculates expected counts, and then to a function which adds columns contaning 
disproportionality.

```{r}
# drug_event_df |>
#   add_expected_counts() |>
#   add_disproportionality()
```

We will now step through these steps in detail. 

### Passing a correctly structured input data
It is important that the report-level data is structured correctly. 
Every line in the passed data frame corresponds to a drug-event pair
from a report.

In the drug_event_df-example data, the first three rows of drug_event_df are from the same report, with report_id = 1. The first row reports a Drug_D and an adverse event 
named Event_5.The next three rows are from another report, where for instance 
drug B has been reported for two different events, event 15 and event 33. 

```{r}
drug_event_df[1:6, ]
```

For completeness, note that if the same drug-event pair occurs several times in 
one report, these drug-event pair contributes to counts only once. That is, 
an observed count of five means that there were five different reports 
containing the drug-event-pair, not e.g. that a single report contained the same 
drug-event pair five times.

### Counting the expected
The details of PRR, RRR and IC are documented within each function, but overall one can note that all expected counts are derived 
from the same contingency table (including marginal sums) of those who did and 
did not have the drug, and the adverse event respectively.

The \code{add_expected_counts}-function has a single parameter, 
\code{expected_count_estimators} that can be used to turn off calculation of 
expected counts for specific estimators. Note that IC by tradition uses 
expected counts referred to as the Relative Reporting Rate (RRR).

```{r}
# drug_event_df |>
#   add_expected_counts(expected_count_estimators = c("prr", "ror", "rrr")) |>
#   head(1)
```


### Adding the disproportionality estimators
The \code{add_disproportionality} function has a \code{da_estimator} parameter, which
by default includes IC, PRR and ROR. If not all of those are wanted, just pass 
the desired estimators instead.

The \code{add_disproportionality} function has two parameters. One is the 
\code{rule_of_N}, by default set to 3, which is sometimes referred to as 
"rule of three". This sets ROR and PRR-values to NA if the observed count is 
less than the specified N. For completeness, note that the default shrinkage 
in the IC acts as a built in 'rule of 3', i.e. the shrinkage of +0.5 prevents 
the lower bound to exceed 0 at the default significance level of 95\%.

The second parameter \code{number_of_digits} controls the rounding of non-count values in the output, 
including all expected counts, uncertainty bounds and point estimates. By default
2 digits are kept. 

```{r}
#| echo=FALSE

# drug_event_df |>
#   add_expected_counts() |>
#   add_disproportionality(
#     rule_of_N = 3,
#     number_of_digits = 2
#   )
```
In short, the add_disproportionality is a wrapper around ror, prr and ic. For 
further details around those estimators, see the documentation of corresponding
function. 

# Subgroup analysis
As the disproportionality builds the comparator from the passed data, 
any subgroup analysis of e.g. specific age groups is achieved by filtering out
the group of interest, and then passing it to `da`.  

# Execution speed 
The data.table-package (through dtplyr) is used for fast execution. A test of 
the execution speed was made using the `vaers` R package, available on [gitlab](https://gitlab.com/iembry/vaers.git). The `vaers` package contains 
data from VAERS from years 1990 - 2018, resulting in 4 146 778 rows to
be processed by `da` in `pvutils`. Execution on a regular laptop took less 
than 7 seconds.

```{r, include=FALSE}

# plot_udfs_as_graph <- function(package_name = "pvutils"){

# library(pvutils)
# package_name <- "pvutils"
# ns_names <- getNamespaceExports(package_name)
# ns_names_w_id <- dplyr::bind_cols("id" = seq_along(ns_names), "name" = ns_names)
# edge_output <- list()
# 
# for (i in seq_along(ns_names)) {
#   # i = 9
# 
#   ns_name_i <- ns_names[i]
#   code_function_i <- body(get(ns_name_i)) |> paste0((as.character()), collapse = "")
# 
#   # Find first occurrence of prefix calls in code_function_i
#   ns_calls <- stringr::str_c(ns_names, "(")
#   # ns_calls_with_comma <- stringr::str_c("", ns_names, "(")
#   # Avoid "is.numeric" being counted as ic
#   code_function_i <- gsub(x = code_function_i, pattern = "as.numeric", replacement = "", fixed = T)
#   # str_replace didn't manage this.
# 
#   first_call_pos <- stringr::str_locate(
#     stringr::fixed(code_function_i),
#     stringr::fixed(c(ns_calls))
#   )[, 1]
# 
#   calls <- na.omit(dplyr::bind_rows(
#     "function_id" = ns_names_w_id$id,
#     "function_name" = ns_names_w_id$name,
#     "pos" = first_call_pos
#   )) |>
#     dplyr::arrange(pos) |>
#     dplyr::select(-pos) |>
#     dplyr::filter(!function_name %in% c("custom_colours", "custom_ggtheme", "sign_lvl_to_quantile_prob"))
# 
#   edges <- dplyr::bind_cols(i, calls) |>
#     dplyr::select(-function_name)
#   colnames(edges) <- c("from", "to")
#   edges <- edges |> 
#     dplyr::filter(from != to)
#   edge_output[[i]] <- edges
# }
# 
# edges <- do.call(rbind.data.frame, edge_output)
# edge_df <- DiagrammeR::create_edge_df(edges$from, edges$to)
# 
# nodes_with_edges <- c(unique(edges$from), unique(edges$to))
# 
# nodes <- dplyr::bind_cols(
#   "id" = 1:length(ns_names),
#   "label" = ns_names
# )[1:length(ns_names) %in% nodes_with_edges, ]
# node_df <- DiagrammeR::create_node_df(length(nodes$id),
#   label = nodes$label,
#   fontsize = 6,
#   shape = "rectangle",
#   fixedsize = FALSE
# )
# 
# node_df$id <- nodes$id
# 
# 
# graph <- DiagrammeR::create_graph(
#   nodes_df = node_df,
#   edges_df = edge_df
# )
# 
# suppressWarnings(DiagrammeR::render_graph(graph,
#   layout = "tree",
#   title = "Overview of package functions"
# ))
```


```{r, echo=FALSE, fig.height = 8, fig.width = 8}

# suppressWarnings(DiagrammeR::render_graph(graph,
#   layout = "tree",
#   title = "Overview of package functions"
# ))
```

