---
title: "Disproportionality_analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Disproportionality_analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

During the life cycle of a medicinal product, knowledge of the safety profile 
continues to accumulate after post-marketing. One way to collect such information
of suspected adverse events is to fill in a spontaneous report and send it to a 
national regulatory authority, such as the FDA, MHRA, PMDA, MPA, etc.      

Given a large set of reports covering a multitude of drugs and adverse events, 
disproportionality analysis is a data driven framework to identify 
drug-event-combinations of potential interest for further qualitative 
manual case-by-case assessment. The pvutils package provides three estimators, 
the proportional reporting rate (PRR), the reporting odds ratio (ROR) and the 
information component (IC). 

## Disproportionality analysis for a specific drug-event combination
A first use-case for these functions is to provide a way to calculate 
confidence or credibility intervals for a specific drug-event-combination, where
you already know the number of observed reports, and other required counts such as
the database total count, number of reports with the exposure and number of
reports with the event.

For instance, you may have extracted counts to an excel sheet from 
some tool where these estimators are not implemented, and want to check 
if the results are consistent across a few different estimators.

```{r}
library(pvutils)
prr(obs = 10, n_drug = 1000, n_event_prr = 200, n_tot_prr = 10000)
ror(a = 10, b = 20, c= 200, d = 10000)
ic(obs = 10, exp = 5)
```
For further details on the specific input parameters, consult the documentation 
of each function. 

## Disproportionality analysis from report-level data
If you do not have the counts mentioned above, but do have report-level data 
that you want to perform disproportionality analysis on, this is straightforward.

Here, we'll demonstrate the use with simulated data set 'drug_event_df' included 
in the package. For simplicity, drugs are named Drug_A - Drug_Z, and events are 
named Event_1 - Event_1000. We pass the data to a function that calculates
expected counts, and then to a function which adds columns contaning 
disproportionality.

```{r}
drug_event_df |>
  add_expected_counts() |>
  add_disproportionality()
```

We will now step through these steps in detail. 

### Passing a correctly structured input data
It is important that the report-level data is structured correctly. 
Every line in the passed data frame corresponds to a drug-event pair
from a report.

In the drug_event_df-example data, the first three rows of drug_event_df are from the same report, with report_id = 1. The first row reports a Drug_D and an adverse event 
named Event_5.The next three rows are from another report, where for instance 
drug B has been reported for two different events, event 15 and event 33. 

```{r}
drug_event_df[1:6, ]
```

For completeness, note that if the same drug-event pair occurs several times in 
one report, these drug-event pair contributes to counts only once. That is, 
an observed count of five means that there were five different reports 
containing the drug-event-pair, not e.g. that a single report contained the same 
drug-event pair five times.

### Counting the expected
Disproportionality analysis is in essence a comparison of two percentages. 
The proportion with an event among reports with an exposure is compared  
to the proportion of the same event in a comparator set of reports, 
often the whole adverse event reports-database.

Inference and output are performed using counts rather than percentages,
in an "observed-to-expected" ratio.  The observed count is compared to an 
expected count, where the latter attempts to provide a baseline reporting 
rate of the event, to highlight drug-event-pairs worthy of further evaluation.

For specific drug-event combinations, there is often potential to construct
more relevant comparators, but disproportionality analysis tries to provide a 
comparator for a wide range of drug-event-pairs, where construction of specific 
comparators would not be feasible. Multiple testing is not an issue, as the
claim is not statistical significance as much as a sorting of potentially 
interesting drug-event pairs. 

The technical details of PRR, RRR and IC are documented within 
each function, but overall one can note that all expected counts are derived 
from the same contingency table (including marginal sums) of those who did and 
did not have the drug, and the adverse event respectively.

The add_expected_counts-function has a single parameter, that can be used
to select if you want to omit expected counts for one of the available 
disproportionality estimators. Note that IC by default uses expected counts 
used in the Relative Reporting Rate (RRR).

```{r}
drug_event_df |> add_expected_counts(da_estimators = c("prr", "ror", "rrr")) |> head(1)
```


### Adding the disproportionality estimators
The add_disproportionality()-function has two parameters. One is the rule_of_N,
by default set to 3, which is sometimes referred to as "rule of three". 
This sets ROR and PRR-values to NA if the observed count is less than the 
passed value. For completeness, note that the default shrinkage in the IC acts as a 
built in 'rule of 3', i.e. the shrinkage of +0.5 prevents the lower bound to 
exceed 0 at the default significance level of 95\%.

The second parameter controls the rounding of non-count values in the output, 
including all expected counts, uncertainty bounds and point estimates. By default
2 digits are kept. 

```{r}
#| echo=FALSE 
  
drug_event_df |> 
  add_expected_counts() |> 
  add_disproportionality(rule_of_N = 3, 
                         number_of_digits = 2)
```
In short, the add_disproportionality is a wrapper around ror, prr and ic. For further details around these parameters, see the documentation of these functions. 

# Subgroup analysis
As the disproportionality builds the comparator from the passed data, 
any subgroup analysis of e.g. specific age groups is achieved by filtering out
the group of interest, and then passing it to add_expected_counts.  

# Technical notes
The package  uses the data.table-package for fast counting, through the dtplyr
wrapper. 

```{r, include=FALSE}

# plot_udfs_as_graph <- function(package_name = "pvutils"){

  library(pvutils)
  package_name = "pvutils"
  ns_names <- getNamespaceExports(package_name)
  ns_names_w_id <- dplyr::bind_cols("id" = seq_along(ns_names), "name"=ns_names)
  edge_output <- list()

  for(i in seq_along(ns_names)){
  # i = 11

  ns_name_i <- ns_names[i]
  code_function_i <- body(get(ns_name_i)) |> paste0((as.character()), collapse="")

  # Find first occurrence of prefix calls in code_function_i
  ns_calls <- stringr::str_c(ns_names, "(")
  # ns_calls_with_comma <- stringr::str_c("", ns_names, "(")


  first_call_pos <- stringr::str_locate(stringr::fixed(code_function_i),
                                  stringr::fixed(c(ns_calls)))[,1]

  calls <- na.omit(dplyr::bind_rows("function_id" = ns_names_w_id$id,
                                    "function_name" = ns_names_w_id$name,
                                    "pos" = first_call_pos)) |>
    dplyr::arrange(pos) |> dplyr::select(-pos) |> 
    dplyr::filter(! function_name %in% c("custom_colours", "custom_ggtheme"))

  edges <- dplyr::bind_cols(i, calls) |> 
    dplyr::select(-function_name)  
  colnames(edges) = c("from", "to")
  edges <- edges |> dplyr::filter(from != to)
  edge_output[[i]] <- edges
  }

  edges <- do.call(rbind.data.frame, edge_output)
  edge_df <- DiagrammeR::create_edge_df(edges$from, edges$to)

  nodes_with_edges <- c(unique(edges$from), unique(edges$to))

  nodes <- dplyr::bind_cols("id" = 1:length(ns_names),
                            "label"=ns_names)[1:length(ns_names) %in% nodes_with_edges,]
  node_df <- DiagrammeR::create_node_df(length(nodes$id),
                            label = nodes$label,
                            fontsize = 6,
                            shape = "rectangle",
                            fixedsize=FALSE)

  node_df$id <- nodes$id


  graph <- DiagrammeR::create_graph(nodes_df = node_df,
                                     edges_df = edge_df)
```


```{r, echo=FALSE, fig.height = 8, fig.width = 8}

suppressWarnings(DiagrammeR::render_graph(graph, 
                                          layout = "tree", 
                                          title="Overview of package functions"))

```

